Ownership types in theory and practice (in Rust)

### 2023-01-23 paper/paper.tex
Added captions to all code snippets in paper

\usepackage{booktabs} % For formal tables
\usepackage{subcaption}
\usepackage{tikz}
\usetikzlibrary{fit}
\usepackage{pgfplots}
\usepackage{pgfplotstable}
\usepackage{hyphenat}
\usepackage{todonotes}
\usepackage[babel]{csquotes}
\usepackage{listings}
\usepackage{listings-rust}

@@

\begin{lstlisting}[language=C++]
std::vector<int> v { 10, 11 };
int *vptr = &v[1]; // points into v
v.push_back(12);
std::cout << *vptr; // bug (use-after-free)
\end{lstlisting}

@@

\begin{lstlisting}[language=Rust]
let mut v = vec![10, 11];
let vptr = &mut v[1]; // points into v
v.push(12);
println!("{}", *vptr); // compiler error
\end{lstlisting}

@@

The objects \emph{car} and \emph{driver} are siblings, meaning they can validly reference each other according to the second rule.

@@

\begin{lstlisting}[language=Rust]
let a = HeapBasedType::default();
let b = a;
\end{lstlisting}
In Rust terminology, \emph{a} is \emph{moved} into \emph{b}.
After the move, \emph{a} is no longer valid, and \emph{b} is the new owner of the value.

Heap-based values will always be generated by a function at runtime.
In the above example, the return value of the function \verb|HeapBasedType::default()| is moved into \emph{a}.
Furthermore, one can generalize the above scenario to ownership being transferred from the variable \emph{a} to the variable \emph{b} when \emph{b} is assigned to the return value \emph{a} of \emph{foo}.
\begin{lstlisting}[language=Rust]
fn main() {
  let b = foo();
}
fn foo() {
  let a = HeapBasedType::default();
  a
}
\end{lstlisting}

Lastly, ownership is transferred from the variable \emph{a} to the variable \emph{b} when \emph{a} is passed to function \emph{foo}, which takes a parameter \emph{b}.
\begin{lstlisting}[language=Rust]
fn main() {
  let a = HeapBasedType::default();
  foo(a);
}
fn foo(b: HeapBasedType) {
  // do something with b
}
\end{lstlisting}
In \emph{foo}, \emph{b} is the owner of the value until it is \emph{dropped} at the end of the function.
Again, accessing the moved-from variable after the ownership of that value has been transferred elsewhere will lead to a compile-time error.
In this case, trying to access \emph{a} after line 3 would not compile, as \emph{a} is no longer the owner of the value.

The heap-allocated \emph{String} type in Rust provides a good example of heap-based ownership transfer.
\begin{lstlisting}[language=Rust]
let x: String = String::from("hello world");
let y: String = x;
println!("x: {}, y: {}", x, y); // error
\end{lstlisting}
After \emph{x} is moved into \emph{y} in line 2, \emph{x} is no longer valid.
Therefore, Rust does not allow accessing the moved-from variable \emph{x} in the third line, and the program will not compile.

@@

The following piece of code \cite{rust-book}, which is similar to the code from above but replaces the \emph{String} type with primitive integers, demonstrates how stack-based values are \emph{copied} instead of \emph{moved} into other variables in Rust.
\begin{lstlisting}[language=Rust]
let x: i32 = 42;
let y: i32 = x;
println!("x: {}, y: {}", x, y);
\end{lstlisting}
The value of \emph{x} is copied into \emph{y}, after which both variables are still valid.
Therefore, the program compiles and "x: 42, y: 42" is printed to stdout.

@@

\begin{lstlisting}[language=Rust]
struct Car {
  engine: Engine,
  wheels: Wheels,
}
fn main() {
  let car = Car::default();
  let moved_engine = car.engine; // move
  car.drive() // error: use of partially moved value: 'car'
}
\end{lstlisting}

@@

The following adjustment to the code snippet from above creates a deep copy of \emph{x} and compiles successfully.
\begin{lstlisting}[language=Rust]
let x: String = String::from("hello world");
let y: String = x.clone();
println!("x: {}, y: {}", x, y);
\end{lstlisting}

@@

For example, an integer can only be reassigned if it is mutable.
\begin{lstlisting}[language=Rust]
let a = 42;
a = 24; // compile error
let mut b = 42;
b = 24; // valid
\end{lstlisting}

@@

\begin{lstlisting}[language=Rust]
fn main() {
  let s1 = String::from("short");
  let s2 = String::from("longest");
  let longest = longest(&s1, &s2);
  println!("longest string: '{}'", longest);
}
fn longest<'a>(s1: &'a str, s2: &'a str) -> &'a str {
  if s1.len() > s2.len() {
    s1
  } else {
    s2
  }
}
\end{lstlisting}
If we omit the lifetime annotations \verb|'a|, the compiler will tell us that the return type of the \verb|longest| function contains a borrowed value, but it is unclear whether it is borrowed from \emph{s1} or \emph{s2}.
In other words, the compiler requires that \emph{s1} and \emph{s2} have to live at least as long as lifetime \verb|'a|.

@@

\begin{lstlisting}[language=Rust]
let s = String::from("hello");
let r1 = &s; // 1. reference
let r2 = &s; // 2. reference
println!("r1: {}, r2: {}", *r1, *r2);
\end{lstlisting}

@@

\begin{lstlisting}[language=Rust]
let mut s = String::from("hello");
String::push_str(&mut s, " world");
// equivalent to
s.push_str(" world");
\end{lstlisting}

@@

\begin{lstlisting}[language=Rust]
let mut s = String::from("hello");

@@

\begin{lstlisting}[language=Rust]
let mut v = vec![10, 11];
let vptr = &mut v[1]; // lifetime 'a starts
Vec::push(&mut v, 12); // lifetime 'b
println!("{}", *vptr); // lifetime 'a ends
\end{lstlisting}

@@

By reordering lines 2 and 3, the lifetimes of the two mutable references will no longer overlap, allowing for a successful compilation of the program.
\begin{lstlisting}[language=Rust]
let mut v = vec![10, 11];
Vec::push(&mut v, 12); // lifetime 'b
let vptr = &mut v[1];  // lifetime 'a starts
println!("{}", *vptr); // lifetime 'a ends
\end{lstlisting}

@@

\begin{lstlisting}[language=C++]
S s{};
S s_copied{s}; // copy constructor
S s_moved{std::move(s)}; // move constructor
// problem: 's' can still be used
\end{lstlisting}

@@

\begin{lstlisting}[language=Rust]
let s = S::default();
let s_copied = s.clone();
let s_moved = s; // 's' is moved
// compiler prevents use of 's'
\end{lstlisting}

>>>>

\usepackage{booktabs} % For formal tables
\usepackage{subcaption}
\usepackage{tikz}
\usetikzlibrary{fit}
\usepackage{pgfplots}
\usepackage{pgfplotstable}
\usepackage{hyphenat}
\usepackage{todonotes}
\usepackage[babel]{csquotes}
\usepackage{listings}
\usepackage{listings-rust}
\usepackage{microtype}

@@

\begin{lstlisting}[language=C++,captionpos=b,caption={C++ use-after-free}]
std::vector<int> v { 10, 11 };
int *vptr = &v[1]; // points into v
v.push_back(12);
std::cout << *vptr; // bug (use-after-free)
\end{lstlisting}

@@

\begin{lstlisting}[language=Rust,captionpos=b,caption={Rust use-after-free}]
let mut v = vec![10, 11];
let vptr = &mut v[1]; // points into v
v.push(12);
println!("{}", *vptr); // compiler error
\end{lstlisting}

@@

The objects \emph{car} and \emph{driver} are siblings because they are both in the ownership context of the object \emph{world}.
This means that they can validly reference each other according to the second rule.

@@

\begin{lstlisting}[language=Rust,captionpos=b,caption={Rust simple heap-based move}]
let a = HeapBasedType::default();
let b = a;
\end{lstlisting}
In Rust terminology, \emph{a} is \emph{moved} into \emph{b}.
After the move, \emph{a} is no longer valid, and \emph{b} is the new owner of the value.

Heap-based values will always be generated by a function at runtime.
In the above example, the return value of the function \verb|HeapBasedType::default()| is moved into \emph{a}.
Furthermore, one can generalize the above scenario to ownership being transferred from the variable \emph{a} to the variable \emph{b} when \emph{b} is assigned to the return value \emph{a} of \emph{foo}.
\begin{lstlisting}[language=Rust,captionpos=b,caption={Rust heap-based move out of function}]
fn main() {
  let b = foo();
}
fn foo() {
  let a = HeapBasedType::default();
  a
}
\end{lstlisting}

Lastly, ownership is transferred from the variable \emph{a} to the variable \emph{b} when \emph{a} is passed to function \emph{foo}, which takes a parameter \emph{b}.
\begin{lstlisting}[language=Rust,captionpos=b,caption={Rust heap-based move into function}]
fn main() {
  let a = HeapBasedType::default();
  foo(a);
}
fn foo(b: HeapBasedType) {
  // do something with b
}
\end{lstlisting}
In \emph{foo}, \emph{b} is the owner of the value until it is \emph{dropped} at the end of the function.
Again, accessing the moved-from variable after the ownership of that value has been transferred elsewhere will lead to a compile-time error.
In this case, trying to access \emph{a} after line 3 would not compile, as \emph{a} is no longer the owner of the value.

The heap-allocated \emph{String} type in Rust provides a good example of heap-based ownership transfer.
\begin{lstlisting}[language=Rust,captionpos=b,caption={Rust String move}]
let x: String = String::from("hello world");
let y: String = x;
println!("x: {}, y: {}", x, y); // error
\end{lstlisting}
After \emph{x} is moved into \emph{y} in line 2, \emph{x} is no longer valid.
Therefore, Rust does not allow accessing the moved-from variable \emph{x} in the third line, and the program will not compile.

@@

The following piece of code \cite{rust-book}, which is similar to the code from above but replaces the \emph{String} type with primitive integers, demonstrates how stack-based values are \emph{copied} instead of \emph{moved} into other variables in Rust.
\begin{lstlisting}[language=Rust,captionpos=b,caption={Rust stack-based move}]
let x: i32 = 42;
let y: i32 = x;
println!("x: {}, y: {}", x, y);
\end{lstlisting}
The value of \emph{x} is copied into \emph{y}, after which both variables are still valid.
Therefore, the program compiles and "x: 42, y: 42" is printed to stdout.

@@

\begin{lstlisting}[language=Rust,captionpos=b,caption={Rust moving struct fields}]
struct Car {
  engine: Engine,
  wheels: Wheels,
}
fn main() {
  let car = Car::default();
  let moved_engine = car.engine; // move
  car.drive() // error: use of partially moved value: 'car'
}
\end{lstlisting}

@@

The following adjustment to the code snippet from above creates a deep copy of \emph{x} and compiles successfully.
\begin{lstlisting}[language=Rust,captionpos=b,caption={Rust copy with clone}]
let x: String = String::from("hello world");
let y: String = x.clone();
println!("x: {}, y: {}", x, y);
\end{lstlisting}

@@

For example, an integer can only be reassigned if it is mutable.
\begin{lstlisting}[language=Rust,captionpos=b,caption={Rust mutable variables}]
let a = 42;
a = 24; // compile error
let mut b = 42;
b = 24; // valid
\end{lstlisting}

@@

\begin{lstlisting}[language=Rust,captionpos=b,caption={Rust lifetime annotations}]
fn main() {
  let s1 = String::from("short");
  let s2 = String::from("longest");
  let longest = longest(&s1, &s2);
  println!("longest string: '{}'", longest);
}
fn longest<'a>(s1: &'a str, s2: &'a str) -> &'a str {
  if s1.len() > s2.len() {
    s1
  } else {
    s2
  }
}
\end{lstlisting}
If we omit the lifetime annotations \verb|'a|, the compiler will tell us that the return type of the \verb|longest| function contains a borrowed value, but it is unclear whether it is borrowed from \emph{s1} or \emph{s2}.
In other words, the compiler requires that \emph{s1} and \emph{s2} have to live at least as long as lifetime \verb|'a|.

@@

\begin{lstlisting}[language=Rust,captionpos=b,caption={Rust immutable references}]
let s = String::from("hello");
let r1 = &s; // 1. reference
let r2 = &s; // 2. reference
println!("r1: {}, r2: {}", *r1, *r2);
\end{lstlisting}

@@

\begin{lstlisting}[language=Rust,captionpos=b,caption={Rust mutable String reference}]
let mut s = String::from("hello");
String::push_str(&mut s, " world");
// equivalent to
s.push_str(" world");
\end{lstlisting}

@@

\begin{lstlisting}[language=Rust,captionpos=b,caption={Rust non-overlapping lifetimes}]
let mut s = String::from("hello");

@@

\begin{lstlisting}[language=Rust,captionpos=b,caption={Rust use-after-free with lifetimes}]
let mut v = vec![10, 11];
let vptr = &mut v[1]; // lifetime 'a starts
Vec::push(&mut v, 12); // lifetime 'b
println!("{}", *vptr); // lifetime 'a ends
\end{lstlisting}

@@

By reordering lines 2 and 3, the lifetimes of the two mutable references will no longer overlap, allowing for a successful compilation of the program.
\begin{lstlisting}[language=Rust,captionpos=b,caption={Rust use-after-free corrected}]
let mut v = vec![10, 11];
Vec::push(&mut v, 12); // lifetime 'b
let vptr = &mut v[1];  // lifetime 'a starts
println!("{}", *vptr); // lifetime 'a ends
\end{lstlisting}

@@

\begin{lstlisting}[language=C++,captionpos=b,caption={C++ move semantics}]
S s{};
S s_copied{s}; // copy constructor
S s_moved{std::move(s)}; // move constructor
// problem: 's' can still be used
\end{lstlisting}

@@

\begin{lstlisting}[language=Rust,captionpos=b,caption={Rust move semantics}]
let s = S::default();
let s_copied = s.clone();
let s_moved = s; // 's' is moved
// compiler prevents use of 's'
\end{lstlisting}

### 2023-01-23 slides/slides.tex
final correction read through of paper
Added captions to all code snippets in paper

\usepackage[utf8]{inputenc}
\usepackage{listings}
\usepackage{listings-rust}
\lstset{
  frame=tb,
  aboveskip=3mm,
  belowskip=3mm,
  showstringspaces=false,
  columns=flexible,
  basicstyle=\footnotesize\ttfamily,
  numbers=left,
  numbersep=5pt,
  numberstyle=\color{gray},
  keywordstyle=\bfseries\color{blue},
  commentstyle=\color{dkgreen},
  stringstyle=\color{mauve},
  breaklines=true,
  breakatwhitespace=false,
  tabsize=2
}
\usepackage{tikz}
\usetikzlibrary{fit,positioning}
\definecolor{tumblue}{RGB}{0,101,189}
\definecolor{dkgreen}{rgb}{0,0.6,0}
\definecolor{mauve}{rgb}{0.58,0,0.82}
\definecolor{cgreen}{RGB}{60,201,65}
\setbeamerfont{footnote}{size=\tiny}
\renewcommand\footnoterule{}
\usecolortheme{orchid}
\setbeamertemplate{navigation symbols}{}
\setbeamertemplate{sidebar right}{% also implies no navbar
  \llap{\tikz\fill[tumblue,scale=.1] (0,0) ++(-5cm,-5cm) ++(-10cm,0)
      +(0cm,0cm) -- +(4cm,0cm) -- +(4cm,-4cm) -- +(5cm,-4cm) -- +(5cm,0cm) --
      +(10cm,0cm) -- +(10cm,-5cm) -- +(9cm,-5cm) -- +(9cm,-1cm) --
      +(8cm,-1cm) -- +(8cm,-5cm) -- +(7cm,-5cm) -- +(7cm,-1cm) --
      +(6cm,-1cm) -- +(6cm,-5cm) -- +(3cm,-5cm) -- +(3cm,-1cm) --
      +(2cm,-1cm) -- +(2cm,-5cm) -- +(1cm,-5cm) -- +(1cm,-1cm) --
      +(0cm,-1cm) -- cycle;}
  % Slide number in triangle.
  \vfill\llap{\tikz\path[fill=tumblue,inner sep=3pt] (0,0) -- ++(-1cm,0) -- +(1cm,1cm) -- cycle node[anchor=south east] {\usebeamerfont{footline}\color{white}\bfseries\insertframenumber};}%
  % Variant without triangle
  % \vfill\llap{\usebeamerfont{footline}\usebeamercolor[fg]{footline}\insertframenumber\hskip3pt}\vskip3pt%
}

@@

\section{Ownership types in theory}

@@

\verb|std::move|:
\begin{itemize}
  \item "move" a named value by casting reference to an R-value reference to tell compiler to use move constructor
  \item no formal effect on the lifetime of the moved-from object $\rightarrow$ moved-from object can still be used, but compiler can't guarantee it is valid
  \item no move constructor or move assignment operator implementation specification
\end{itemize}
\end{frame}

@@

\section{Conclusions and outlook}

@@

\section{Extra slides}

\begin{frame}{Lifetime elision}
Here are the three rules:
\begin{itemize}
  \item Each elided lifetime in a function’s arguments becomes a distinct lifetime parameter.
  \item If there is exactly one input lifetime, elided or not, that lifetime is assigned to all elided lifetimes in the return values of that function.
  \item If there are multiple input lifetimes, but one of them is &self or &mut self, the lifetime of self is assigned to all elided output lifetimes.
\end{itemize}
Otherwise, it is an error to elide an output lifetime.
\end{frame}

\begin{frame}[fragile]{Lifetime elision: examples}
\begin{lstlisting}[language=Rust]
fn print(s: &str); // elided
fn print<'a>(s: &'a str); // expanded

fn debug(lvl: u32, s: &str); // elided
fn debug<'a>(lvl: u32, s: &'a str); // expanded

fn substr(s: &str, until: u32) -> &str; // elided
fn substr<'a>(s: &'a str, until: u32) -> &'a str; // expanded

fn get_str() -> &str; // ILLEGAL, no inputs

fn frob(s: &str, t: &str) -> &str; // ILLEGAL, two inputs
fn frob<'a, 'b>(s: &'a str, t: &'b str) -> &str; // Expanded: Output lifetime is ambiguous
\end{lstlisting}
\end{frame}

>>>>

\usepackage[utf8]{inputenc}
\usepackage{listings}
\usepackage{listings-rust}
\lstset{
  frame=tb,
  aboveskip=3mm,
  belowskip=3mm,
  showstringspaces=false,
  columns=flexible,
  basicstyle=\footnotesize\ttfamily,
  numbers=left,
  numbersep=5pt,
  numberstyle=\color{gray},
  keywordstyle=\bfseries\color{blue},
  commentstyle=\color{dkgreen},
  stringstyle=\color{mauve},
  breaklines=true,
  breakatwhitespace=false,
  tabsize=2
}
\usepackage{tikz}
\usetikzlibrary{fit,positioning}
\definecolor{tumblue}{RGB}{0,101,189}
\definecolor{dkgreen}{rgb}{0,0.6,0}
\definecolor{mauve}{rgb}{0.58,0,0.82}
\definecolor{cgreen}{RGB}{60,201,65}
\setbeamerfont{footnote}{size=\tiny}
\renewcommand\footnoterule{}
\usecolortheme{orchid}
\setbeamertemplate{navigation symbols}{}
\setbeamertemplate{sidebar right}{% also implies no navbar
  \llap{\tikz\fill[tumblue,scale=.1] (0,0) ++(-5cm,-5cm) ++(-10cm,0)
      +(0cm,0cm) -- +(4cm,0cm) -- +(4cm,-4cm) -- +(5cm,-4cm) -- +(5cm,0cm) --
      +(10cm,0cm) -- +(10cm,-5cm) -- +(9cm,-5cm) -- +(9cm,-1cm) --
      +(8cm,-1cm) -- +(8cm,-5cm) -- +(7cm,-5cm) -- +(7cm,-1cm) --
      +(6cm,-1cm) -- +(6cm,-5cm) -- +(3cm,-5cm) -- +(3cm,-1cm) --
      +(2cm,-1cm) -- +(2cm,-5cm) -- +(1cm,-5cm) -- +(1cm,-1cm) --
      +(0cm,-1cm) -- cycle;}
  % Slide number in triangle.
  \vfill\llap{\tikz\path[fill=tumblue,inner sep=3pt] (0,0) -- ++(-1cm,0) -- +(1cm,1cm) -- cycle node[anchor=south east] {\usebeamerfont{footline}\color{white}\bfseries\insertframenumber};}%
  % Variant without triangle
  % \vfill\llap{\usebeamerfont{footline}\usebeamercolor[fg]{footline}\insertframenumber\hskip3pt}\vskip3pt%
}
% TODO: navigation bar
% \setbeamertemplate{footline}{%
%  \begin{beamercolorbox}[ignorebg,ht=2.25ex,dp=3.75ex]%
%        {section in head/foot}%
%    \insertsectionnavigationhorizontal{\dimexpr\paperwidth-1cm}{}{}%
%  \end{beamercolorbox}%
% }

@@

\section{Ownership types}

@@

\verb|std::move|:
\begin{itemize}
  \item "move" a named value by casting it to an R-value reference to tell compiler to use move constructor
  \item no formal effect on the lifetime of the moved-from object $\rightarrow$ moved-from object can still be used, but compiler can't guarantee it is valid
  \item no move constructor or move assignment operator implementation specification
\end{itemize}
\end{frame}

@@

\section{Conclusion}

@@

% \section{Extra slides}

% \begin{frame}{Lifetime elision}
% Here are the three rules:
% \begin{itemize}
%   \item Each elided lifetime in a function’s arguments becomes a distinct lifetime parameter.
%   \item If there is exactly one input lifetime, elided or not, that lifetime is assigned to all elided lifetimes in the return values of that function.
%   \item If there are multiple input lifetimes, but one of them is &self or &mut self, the lifetime of self is assigned to all elided output lifetimes.
% \end{itemize}
% Otherwise, it is an error to elide an output lifetime.
% \end{frame}

% \begin{frame}[fragile]{Lifetime elision: examples}
% \begin{lstlisting}[language=Rust]
% fn print(s: &str); // elided
% fn print<'a>(s: &'a str); // expanded

% fn debug(lvl: u32, s: &str); // elided
% fn debug<'a>(lvl: u32, s: &'a str); // expanded

% fn substr(s: &str, until: u32) -> &str; // elided
% fn substr<'a>(s: &'a str, until: u32) -> &'a str; // expanded

% fn get_str() -> &str; // ILLEGAL, no inputs

% fn frob(s: &str, t: &str) -> &str; // ILLEGAL, two inputs
% fn frob<'a, 'b>(s: &'a str, t: &'b str) -> &str; // Expanded: Output lifetime is ambiguous
% \end{lstlisting}
% \end{frame}
